---
title: "R_in_action. Simple Health Data Analysis: Logistic Regression"
output: html_document
date: "2026-09-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

## Dataset: HANES_R_Ignite.csv (NHANES, 10,000 observations)
## Outcome: Diabetes (Yes/No)

# 1. Load Data
```{r}
df <- read.csv("HANES_R_Ignite.csv", stringsAsFactors = TRUE)
cat("Rows:", nrow(df), " Columns:", ncol(df), "\n")
```

# 2. Select relevant variables & clean

```{r}
vars <- c("Diabetes", "Age", "Gender", "BMI", "BPSysAve", "BPDiaAve",
          "TotChol", "PhysActive", "Smoke100")
df <- df[, vars]
df <- na.omit(df)
cat("Rows after removing missing values:", nrow(df), "\n")
```

# 3. Quick EDA
```{r}
cat("\n--- Summary ---\n")
print(summary(df))
```
# 4. Train/test split (70/30)
```{r}
set.seed(42)
n <- nrow(df)
train_idx <- sample(1:n, size = 0.7 * n)
train <- df[train_idx, ]
test  <- df[-train_idx, ]
```
# 5. Logistic regression model
```{r}
model <- glm(Diabetes ~ Age + Gender + BMI + BPSysAve + BPDiaAve +
               TotChol + PhysActive + Smoke100,
             data = train, family = binomial)

cat("\n--- Logistic Regression Summary ---\n")
print(summary(model))

cat("\n--- Odds Ratios ---\n")
print(round(exp(coef(model)), 3))
```
# 6. Predict on test set
```{r}
pred_prob <- predict(model, newdata = test, type = "response")
pred_class <- ifelse(pred_prob > 0.5, "Yes", "No")
```
# 7. Evaluate
```{r}
accuracy <- mean(pred_class == test$Diabetes)
cat("\n--- Model Evaluation ---\n")
cat("Accuracy:", round(accuracy, 3), "\n")

cat("\nConfusion Matrix:\n")
print(table(Predicted = pred_class, Actual = test$Diabetes))
```
## =========================================================
## Simple Health Data Analysis: Random Forest (Classical ML)
## Dataset: HANES_R_Ignite.csv (NHANES, 10,000 observations)
## Outcome: Diabetes (Yes/No)
## =========================================================
install.packages("randomForest")
library(randomForest)

# 1. Load Data
```{r}
df <- read.csv("HANES_R_Ignite.csv", stringsAsFactors = TRUE)
cat("Rows:", nrow(df), " Columns:", ncol(df), "\n")
```
# 2. Select relevant variables & clean
```{r}
vars <- c("Diabetes", "Age", "Gender", "BMI", "BPSysAve", "BPDiaAve",
          "TotChol", "PhysActive", "Smoke100")
df <- df[, vars]
df <- na.omit(df)
df$Diabetes <- factor(df$Diabetes)  # ensure classification, not regression
cat("Rows after removing missing values:", nrow(df), "\n")
```
# 3. Quick EDA
```{r}
cat("\n--- Diabetes counts ---\n")
print(table(df$Diabetes))
```
# 4. Train/Test split (70/30)
```{r}
set.seed(42)
n <- nrow(df)
train_idx <- sample(1:n, size = 0.7 * n)
train <- df[train_idx, ]
test  <- df[-train_idx, ]
```
# 5. Random Forest model
```{r}
rf_model <- randomForest(Diabetes ~ Age + Gender + BMI + BPSysAve + BPDiaAve +
                            TotChol + PhysActive + Smoke100,
                          data = train, ntree = 500, importance = TRUE)
 
cat("\n--- Random Forest Summary ---\n")
print(rf_model)
 
cat("\n--- Variable Importance ---\n")
print(importance(rf_model))
```
# 6. Predict on test set
```{r}
pred_class <- predict(rf_model, newdata = test)
```
# 7. Evaluate
```{r}
accuracy <- mean(pred_class == test$Diabetes)
cat("\n--- Model Evaluation ---\n")
cat("Accuracy:", round(accuracy, 3), "\n")
 
cat("\nConfusion Matrix:\n")
print(table(Predicted = pred_class, Actual = test$Diabetes))
```
# 8. -- Additional metrics ---
install.packages("pROC")
# Get predicted probabilites (needed for ROC/AUC)
```{r}
pred_prob <- predict(rf_model, newdata = test, type = "prob")[, "Yes"]
```
# Precision, Recall, F1 (from the confusion matrix)
```{r}
cm <- table(Predicted = pred_class, Actual = test$Diabetes)
TP <- cm["Yes", "Yes"]
FP <- cm["Yes", "No"]
FN <- cm["No", "Yes"]
precision <- TP / (TP + FP)
recall    <- TP / (TP + FN)
f1        <- 2 * precision * recall / (precision + recall)

cat("Precision:", round(precision, 3), "\n")
cat("Recall:", round(recall, 3), "\n")
cat("F1 Score:", round(f1, 3), "\n")

# ROC curve & AUC (uses pROC package)
library(pROC)
roc_obj <- roc(test$Diabetes, pred_prob, levels = c("No", "Yes"), quiet = TRUE)
cat("AUC:", round(auc(roc_obj), 3), "\n")
```
# --- 1. Variable importance plot (ggplot2) ---
```{r}
imp_df <- as.data.frame(importance(rf_model))
imp_df$Variable <- rownames(imp_df)
library(ggplot2)
ggplot(imp_df, aes(x = reorder(Variable, MeanDecreaseAccuracy),
                    y = MeanDecreaseAccuracy)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Random Forest Variable Importance",
       x = "Variable", y = "Mean Decrease in Accuracy") +
  theme_minimal()
```
# --- 2. ROC curve (ggplot2) ---
```{r}
roc_df <- data.frame(
  specificity = rev(roc_obj$specificities),
  sensitivity = rev(roc_obj$sensitivities)
)

ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_abline(linetype = "dashed", color = "grey50") +
  labs(title = paste("ROC Curve (AUC =", round(auc(roc_obj), 3), ")"),
       x = "1 - Specificity", y = "Sensitivity") +
  theme_minimal()
```

## --- 3. Error rate vs number of trees (ggplot2) ---
```{r}
err_df <- as.data.frame(rf_model$err.rate)
err_df$Trees <- 1:nrow(err_df)
err_long <- reshape(err_df, varying = names(err_df)[1:3],
                     v.names = "ErrorRate", timevar = "Type",
                     times = names(err_df)[1:3], direction = "long")

ggplot(err_long, aes(x = Trees, y = ErrorRate, color = Type)) +
  geom_line(linewidth = 0.8) +
  labs(title = "RF Error Rate vs Number of Trees",
       x = "Number of Trees", y = "Error Rate") +
  theme_minimal()
```

## --- 4. Predicted probability by actual class (ggplot2) ---
```{r}
plot_df <- data.frame(pred_prob = pred_prob, Actual = test$Diabetes)

ggplot(plot_df, aes(x = Actual, y = pred_prob, fill = Actual)) +
  geom_boxplot() +
  labs(title = "Predicted Probability by Actual Diabetes Status",
       x = "Actual", y = "Predicted P(Diabetes = Yes)") +
  theme_minimal() +
  theme(legend.position = "none")
```